Parsing Postfix log files

### 2007-06-18 doc/logparser.tex
Wrote the SMTP background section, plus supporting references.  Started on parsing efficiency, but didn't get far.  Added Log Mail Analyser to the list of reviewed parsers.  Improved one graph slightly.

XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS

@@

@@

@@

>>>>

XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS.  REVIEW
EACH PAPER, WRITING AN ABSTRACT FOR IT: WHAT'S THE SAME, WHAT'S DIFFERENT,
PERFORMANCE AND \newline COMPLETENESS COMPARISON.  FROM THIS I'LL GET AN
OVERALL PICTURE, AND I CAN WORK REFERENCES TO THE PAPERS INTO THIS PAPER,
PLUS MAYBE A PRIOR ART SECTION.

XXX TURN OFF RESTRICTIONS FOR ME AND SEE HOW WELL SPAMASSASSIN DEALS
WITH THE RESULTING MAILS; FALSE POSITIVES, FALSE NEGATIVES, ETC.

@@

The Simple Mail Transfer Protocol, originally defined in RFC
821~\cite{RFC821} and later in RFC 2821~\cite{RFC2821}, is used for
transferring mail between Mail Transport Agents (MTAs).  It is a simple,
human readable, plain text protocol, making it quite easy to test and debug
problems with it.  Despite the simplicity many virus and/or spam sending
programs fail to implement it properly, so requiring strict adherence to
the protocol specification is beneficial in protecting against spam and
viruses.\footnote{This violates the principle of \textit{Be liberal in what
you accept, and conservative in what you send\/} from RFC
760~\cite{rfc760}, but unfortunately that principle was written in a
friendlier time.}  A typical SMTP conversation resembles the following (the
lines starting with a three digit number are sent by the server, all other
lines are sent by the client):

\begin{verbatim}
220 smtp.example.com ESMTP
HELO client.example.com
250 smtp.example.com
MAIL FROM: <alice@example.com>
250 2.1.0 Ok
RCPT TO: <bob@example.com>
250 2.1.5 Ok
DATA
354 End data with <CR><LF>.<CR><LF>
Message headers and body sent here.
.
250 2.0.0 Ok: queued as D7AFA38BA
QUIT
221 2.0.0 Bye
\end{verbatim}

A detailed description of SMTP is beyond the scope of this document:
introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 50 MB log files containing 300,000 log lines.  The graphs in
section~\ref{graphs} illustrate the execution time for various
configurations of the parser and should be referred to during the following
discussion.

An important property of a parser is how execution time scales with input
size: does it scale linearly, polynomially, or exponentially?
Graph~\ref{execution time vs file size vs number lines} shows the execution
time in seconds, file size in MB and tens of thousands of lines per log
file.  All three lines run roughly parallel, giving a visual impression
that the algorithm scales linearly with input size.  This is borne(XXX) out
by graph~\ref{execution time vs file size vs number lines factor}, which
plots the ratio of file size in MB vs execution time in seconds and number
of lines

@@

    \item [Log Mail Analyser] \cite{log-mail-analyser} attempts to
        correlate Postfix log lines by queueid and produce both a CSV file
        and either a Berkeley DB or MySQL database.  It aims to save
        \textit{date and hour, DNS name and IP address host, mail server IP
        address, sender, receiver and e-mail status (sent, rejected)\/}, so
        its data extraction is reasonably complete.  The parsing is
        hard-coded into the program; extending it requires familiarity with
        a substantial portion of the code.  Producing a CSV file appears to
        be quite limiting, but the intention is to provide a format
        suitable for use with grep(1), awk(1) and the myriad of other Unix
        utilities already available.  Supporting Berkeley DB is also
        limiting, requiring an understanding of the (partially documented)
        database structure and writing a program to extract the required
        data - it appears to be the least useful of the three output
        formats, though the authors of LMA did provide some queries:
        IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT.  Unfortunately
        when the author tested LMA it didn't work properly and after
        exploring the code for some time the author reluctantly gave up.
